---
title: "Class 17"
author: "Kimberly Navarro"
date: "2026-03-04"
format: pdf
---

## Downstream Analysis

Download "txtimport" and "rhdf5" in terminal using BiocManager.

BiocManager::install("tximport") & 
BiocManager::install("rhdf5")

```{r}
library(tximport)

folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.h5" )
names(files) <- samples
txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```
Estimate transcript counts per sample:

```{r}
head(txi.kallisto$counts)
```

Transcripts per sample: 

```{r}
colSums(txi.kallisto$counts)
```

Transcripts detected in at least one sample: 

```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

Filter out annotated transcripts with no reads: 

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
```

Filter out those with no change over the samples:

```{r}
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```

## Principal Component Analysis

Now we compute the principal components, centering and scaling each transcript’s measured levels so that each feature contributes equally to the PCA:

```{r}
pca <- prcomp(t(x), scale=TRUE)
```

```{r}
summary(pca)
```

```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```
>Q. Use ggplot to make a similar figure of PC1 vs PC2 and a separate figure PC1 vs PC3 and PC2 vs PC3.

PC1 vs PC2 plot:

```{r}
library(ggplot2)
library(ggrepel)

mycols <- c("blue","blue","red","red")

ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point( col=mycols ) +
  geom_text_repel( col=mycols ) +
  theme_bw()
```

PC1 vs PC3 plot:

```{r}
library(ggplot2)
library(ggrepel)

mycols <- c("blue","blue","red","red")

ggplot(pca$x) +
  aes(PC1, PC3, label=rownames(pca$x)) +
  geom_point( col=mycols ) +
  geom_text_repel( col=mycols ) +
  theme_bw()
```

PC2 vs PC3 plot:

```{r}
library(ggplot2)
library(ggrepel)

mycols <- c("blue","blue","red","red")

ggplot(pca$x) +
  aes(PC2, PC3, label=rownames(pca$x)) +
  geom_point( col=mycols ) +
  geom_text_repel( col=mycols ) +
  theme_bw()
```